Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
pass through 8.1

Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool} for each file of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the extension to
normalization for open and closed terms from Section~\ref{sec:extension}.

@@

\def\rot{\rotatebox}
\newcommand\header[1]{\rotatebox{90}{{#1}}}

@@

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{ l |  l  | l | l | l }
      & \header{Universes} & \header{Inductives} & \header{Conversion}
      & \header{Large Elim} \\
    \hline
    \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
    \hline
    CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    CoreSpec & \\
    PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}
% \end{minipage}

@@

\subsection{Logical relations}
Proof by logical relations does not come with a precise definition. In
the most general sense, a logical relation can be viewed as a
practical technique that uses a type-indexed relation to assign
meanings to types and strengthen the induction hypothesis for the
property of interest. The original idea of this technique can be
traced back to
\citet{tait1967:reducibility}. \citet{tait1967:reducibility} maps
types to sets of terms satisfying certain properties related to reduction.
The same idea is explained in \citet{girard1989proofs} and extended to
prove the strong normalization property of System F.

Tait's method has been successfully applied to dependently typed
languages.  \citet{Martin-Lof-1973}, \citet{luo1990extended},
\citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that prove strong
normalization for different dependently type systems using Tait's notion of
reducibility.

\scw{Should give Martin L\"of credit for first consistency proof.}
% \scw{What techniques do these three proofs use? How is it different from
% your proof? What about Luo, which
% introduced a universe hierarchy? What about categories with families, or
% other ways of showing consistency?}
% \yl{resolved}

% Some of these proofs are still reapplied in more recent works. For
% example, the technique for proving strong normalization by
% \citet{geuvers1994short} is adapted by \citet{moon2021graded} to show
% the normalization property for a dependently typed system extended with
% modalities.

The pen and paper representation of the logical relation proofs
can be challenging to adapt to a theorem prover since a lot of details
are hidden behind the concise notations.
As an example, \citet{geuvers1994short} presents the interpretation for types as
an inductively defined total function over the set of syntactically
well-formed types. Despite the notation for the logical relation as a
simply typed function that takes a type and returns some set, the
interpretation function in reality is a dependent function whose
return type depends on the derivation of the well-typedness of its
input. The well-typedness derivation and the proof of the
classification theorem are needed relevantly in the body
of the interpretation function to decide whether an argument of an
application should be erased during interpretation. Due to the
impredicativity of the object language, \citet{geuvers1994short}'s
proof cannot be encoded in Agda, which has a predicative
metatheory. Due to the use of proof-relevant derivations, even in
Coq, a proof assistant that supports impredicativity, one would need
to constantly juggle between the impredicative but irrelevant sort
\texttt{Prop} sort and the predicative but relevant sort
\texttt{Type}.

More recent works such as \citet{Abel12}
and\citet{abel2008betaeta} make their definitions more explicit and
precise and thus easier to encode in proof assistants. Our definition
of logical relation is reminiscent of their definition of a
semantic universe hierarchy, though our logical relation is defined to
be closed under expansion with respect to parallel reduction rather
than weak head reduction. Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use the semantic universe hierarchy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly in our definition of
semantic typing because it is sufficient
for deriving consistency and later normalization.

\subsection{Mechanized logical relations about dependent types}
In Figure~\ref{fig:featurematrix}, we give an overview of features of
the object languages from mechanized logical relations about dependent
types. The table is far from exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.

>>>>

Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool, available from \url{https://github.com/AlDanial/cloc}.} for each file of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the extension to
normalization for open and closed terms from Section~\ref{sec:extension}.

@@

\subsection{Logical relations for dependent types}
In the most general sense, a logical relation can be viewed as a
practical technique that uses a type-indexed relation to assign
meanings to types and strengthen the induction hypothesis for the
property of interest. The original idea of this technique can be
traced back to
\citet{tait1967:reducibility}. This proof maps
types to sets of terms satisfying certain properties related to reduction.
The same idea is explained in \citet{girard1989proofs} and extended to
prove the strong normalization property of System F.

Tait's method has been successfully applied to dependently typed
languages to prove strong normalization \cite{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.

However, the pen-and-paper representation of logical relations proofs
can be challenging to adapt to a theorem prover since many details
are hidden behind concise notations.
For example, \citet{geuvers1994short} presents the interpretation for types as
an inductively defined total function over the set of syntactically
well-formed types. \scw{The issue is that Geuver's metalogic is (untyped) set 
theory, but Coq and Agda use type theory.}
In untyped set theory, it makes sense to define the logical relation as a
simply-typed function that takes a type and returns some set; however
in constructive type theory, the metalogic of Coq and Agda, the
interpretation function must be a dependently-typed function whose
return type depends on the derivation of the well-typedness of its
input. The well-typedness derivation and the proof of the
classification theorem\scw{what is the classification theorem}
are examined in the body
of the interpretation function to decide whether an argument of an
application should be erased during interpretation. As a result, this 
definition causes difficulties for modern proof assistants.
Due to the
impredicativity of the object language, \citet{geuvers1994short}'s
proof cannot be encoded in Agda, which has a predicative
metatheory. Due to the use of proof-relevant derivations, even in
Coq, a proof assistant that supports impredicativity, one would need
to constantly juggle between the impredicative but irrelevant sort
\texttt{Prop} sort and the predicative but relevant sort
\texttt{Type}.

More recent work such as \citet{Abel12}
and \citet{abel2008betaeta} make their definitions more explicit and
precise and thus more directly encodable in proof assistants. 
Our logical relation resembles their definition of a
semantic universe hierarchy, although we close our relation 
under expansion with respect to parallel reduction rather
than weak-head reduction. \scw{Why is this important?}
Furthermore, \citet{Abel12} and
\citet{abel2008betaeta} use their semantic universe hierarchy as a
measure to define Kripke-style logical relations, from which they
derive the correctness of their conversion algorithms. In our work, we
use the semantic universe hierarchy directly in our definition of
semantic typing because it is sufficient for our purposes
(consistency and normalization).

\subsection{Mechanized logical relations for dependent types}

\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
\newcommand\header[1]{\rotatebox{90}{{#1}}}

@@

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{ l |  l  | l | l | l }
      & \header{Universes} & \header{Inductives} & \header{Conversion}
      & \header{Large Elim} \\
    \hline
    \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
    % \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}
% \end{minipage}

@@

In Figure~\ref{fig:featurematrix}, we give an overview of features of
the object languages from mechanized logical relations about dependent
types. The table is far from exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.
